Raw Data Comparison

### 2019-09-13 Raw Data Comparison.Rmd
...

```{r, echo=F, message=F, warning=F}
files = list.files('~/Data/BarrettsProgressionRisk/Analysis/multipcf_perPatient/', '5e06_cleaned_tiled', recursive=T, full.names = T)
training.tiles = do.call(bind_rows, purrr::map(files, function(f) {
  read_tsv(f,col_types=c(.default=col_double()))
})) %>% dplyr::rename('Sample' = 'X1')

@@

## Per patient

Per patient after tiling which merges the segments using a weighted mean.  Again, there's an upwards shift per-patient. 
```{r, fig.width=10}

@@

gridExtra::grid.arrange(
  ggplot(bind_rows(t.pt, v.pt), aes(cohort, mean, group=cohort, fill=cohort)) + 
    geom_jitter(width=0.1) + geom_violin(alpha=0.5) + theme(legend.position = 'none') +
    labs(title='Mean per patient'),
  ggplot(bind_rows(t.pt, v.pt), aes(cohort, sd, group=cohort, fill=cohort)) + 
    geom_jitter(width=0.1) + geom_violin(alpha=0.5) + theme(legend.position = 'none') +
    labs(title='StdDev per patient'),
  ncol=2, top='Per patient tiled values')

t.per.pt.scale = as_tibble(t(apply(as.matrix(training.tiles[,-1]), 1, scale, center=T, scale=T))) %>% add_column('sample'=training.tiles$Sample, .before=1)
colnames(t.per.pt.scale) = colnames(training.tiles)
t.arm.pp.scale = as_tibble(t(apply(as.matrix(training.arms[,-1]), 1, scale, center=T, scale=T))) %>% add_column('sample'=training.arms$Sample, .before=1)
colnames(t.arm.pp.scale) = colnames(training.arms)

@@

## Per bin

Per 5MB bin (across patients).  There's a lot more variation in the means per bin in the validation data. This is before unit normalization, which is based on the training data mean/sd, so where the variance causes issues is starting to become more clear.

```{r}
tile.dist.measure<-function(df) {
  means = as_tibble(df %>% dplyr::summarise_at(vars(-Sample), list(~mean(.,na.rm=T))) %>% t, rownames='bin')  %>% dplyr::rename('mean' = 'V1')
  sds = as_tibble(df %>% dplyr::summarise_at(vars(-Sample), list(~sd(.,na.rm=T))) %>% t, rownames='bin')  %>% dplyr::rename('sd' = 'V1')
  iqrs = as_tibble(df %>% dplyr::summarise_at(vars(-Sample), list(~IQR(.,na.rm=T))) %>% t, rownames='bin')  %>% dplyr::rename('iqrs' = 'V1')
  vars = as_tibble(df %>% dplyr::summarise_at(vars(-Sample), list(~var(.,na.rm=T))) %>% t, rownames='bin')  %>% dplyr::rename('var' = 'V1')

t.dist = tile.dist.measure(training.tiles) %>% mutate(cohort='train')
v.dist = tile.dist.measure(val.tiles) %>% mutate(cohort='val')

@@

### Per bin, after scaling per-pt

gridExtra::grid.arrange(
  ggplot(bind_rows(t.dist, v.dist), aes(cohort, mean, group=cohort, fill=cohort)) + 
    geom_jitter(width=0.1) + geom_violin(alpha=0.5) + theme(legend.position = 'none') +
    labs(title='Mean'),
  ggplot(bind_rows(t.dist, v.dist), aes(cohort, sd, group=cohort, fill=cohort)) + 
    geom_jitter(width=0.1) + geom_violin(alpha=0.5) + theme(legend.position = 'none') +
    labs(title='SD'), 
  ncol=2, top='Per bin tiled values on per-pt scaled')
```

After the arm segments have been subtracted as well.
```{r}

## Normalize validation cohort within itself

Instead scaling using z-values from the trianing data, I can do it from within the validation cohort. This looks better here, but when I use this for predictions it doesn't significantly change the results.

@@

@@

# Non progressors
preds %>% dplyr::filter(Status == 'NP') %>% group_by(Risk) %>% tally() %>% spread(Risk, n)

# FPR is nearly 40% still
nrow(preds %>% filter(Status == 'NP' & Risk == 'High'))/nrow(preds %>% filter(Status == 'NP'))

# Progressors
preds %>% dplyr::filter(Status == 'P' & !grepl('OCCAMS',`Hospital Research ID`)) %>% group_by(Risk) %>% tally() %>% spread(Risk, n)

# FNR is still over 50% in the old progressors as well
nrow(preds %>% filter(Status == 'P' & Risk == 'Low' & !grepl('OCCAMS',`Hospital Research ID`)))/nrow(preds %>% filter(Status == 'P' &  !grepl('OCCAMS',`Hospital Research ID`)))
```
```

>>>>

```{r, echo=F, message=F, warning=F}
tile.dist.measure<-function(df) {
  means = as_tibble(df %>% dplyr::summarise_at(vars(-Sample), list(~mean(.,na.rm=T))) %>% t, rownames='bin')  %>% dplyr::rename('mean' = 'V1')
  sds = as_tibble(df %>% dplyr::summarise_at(vars(-Sample), list(~sd(.,na.rm=T))) %>% t, rownames='bin')  %>% dplyr::rename('sd' = 'V1')
  iqrs = as_tibble(df %>% dplyr::summarise_at(vars(-Sample), list(~IQR(.,na.rm=T))) %>% t, rownames='bin')  %>% dplyr::rename('iqrs' = 'V1')
  vars = as_tibble(df %>% dplyr::summarise_at(vars(-Sample), list(~var(.,na.rm=T))) %>% t, rownames='bin')  %>% dplyr::rename('var' = 'V1')

files = list.files('~/Data/BarrettsProgressionRisk/Analysis/multipcf_perPatient/', '5e06_cleaned_tiled', recursive=T, full.names = T)
training.tiles = do.call(bind_rows, purrr::map(files, function(f) {
  read_tsv(f,col_types=c(.default=col_double()))
})) %>% dplyr::rename('Sample' = 'X1')

@@

## Looking at the distribution of values per sample

Per sample after tiling which merges the segments using a weighted mean.  Again, there's an upwards shift per-sample 
```{r, fig.width=10}

@@

gridExtra::grid.arrange(
  ggplot(bind_rows(t.pt, v.pt), aes(cohort, mean, group=cohort, fill=cohort)) + 
    geom_jitter(width=0.1) + geom_violin(alpha=0.5) + theme(legend.position = 'none') +
    labs(title='Mean per patient'),
  ggplot(bind_rows(t.pt, v.pt), aes(cohort, sd, group=cohort, fill=cohort)) + 
    geom_jitter(width=0.1) + geom_violin(alpha=0.5) + theme(legend.position = 'none') +
    labs(title='StdDev per patient'),
  ncol=2, top='Per patient tiled values')
```

## Scale per sample

Scaling and centering per sample makes the samples directly comparable, but not sure what it will do to the bins.  
```{r}
t.per.pt.scale = as_tibble(t(apply(as.matrix(training.tiles[,-1]), 1, scale, center=T, scale=T))) %>% add_column('sample'=training.tiles$Sample, .before=1)
colnames(t.per.pt.scale) = colnames(training.tiles)
t.arm.pp.scale = as_tibble(t(apply(as.matrix(training.arms[,-1]), 1, scale, center=T, scale=T))) %>% add_column('sample'=training.arms$Sample, .before=1)
colnames(t.arm.pp.scale) = colnames(training.arms)

@@

### Per bin, after scaling per-sample

Now the variance of the cohorts appears to be nearly the same.

Just to check...after the arm segments have been subtracted as well.
```{r}

gridExtra::grid.arrange(
  ggplot(bind_rows(t.dist, v.dist), aes(cohort, mean, group=cohort, fill=cohort)) + 
    geom_jitter(width=0.1) + geom_violin(alpha=0.5) + theme(legend.position = 'none') +
    labs(title='Mean'),
  ggplot(bind_rows(t.dist, v.dist), aes(cohort, sd, group=cohort, fill=cohort)) + 
    geom_jitter(width=0.1) + geom_violin(alpha=0.5) + theme(legend.position = 'none') +
    labs(title='SD'), 
  ncol=2, top='Per bin tiled values on per-sample scaled')

Using a model built on per-sample scaled data. Unfortunately that has not resolved the predictions.

## Per bin variance

No scaling at this point.

Per 5MB bin (across patients).  There's a lot more variation in the means per bin in the validation data. This is before bin scaling, which is based on the training data mean/sd, so where the variance causes issues is starting to become more clear.

```{r}
t.dist = tile.dist.measure(training.tiles) %>% mutate(cohort='train')
v.dist = tile.dist.measure(val.tiles) %>% mutate(cohort='val')

@@

## Scale validation cohort within itself

Instead scaling using z-values from the training data, I can do it from within the validation cohort. This looks better here, but when I use this for predictions it doesn't significantly change the results.

@@

## Scale validation using global z scores

```{r}
v.means = apply(as.matrix(bind_rows(training.tiles, val.tiles)[,-1]),2, mean, na.rm=T)
v.sd = apply(as.matrix(bind_rows(training.tiles, val.tiles)[,-1]),2, sd, na.rm=T)

val.int.norm = as_tibble(val.int.norm) %>% add_column('Sample'=val.tiles$Sample,.before=T)
vinorm.dist = tile.dist.measure(val.int.norm) %>% mutate(cohort='val glob')

gridExtra::grid.arrange(
  ggplot(bind_rows(tnorm.dist, vinorm.dist), aes(cohort, mean, group=cohort, fill=cohort)) + 
    geom_jitter(width=0.1) + geom_violin(alpha=0.5) + theme(legend.position = 'none') +
    labs(title='Mean'),
  ggplot(bind_rows(tnorm.dist, vinorm.dist), aes(cohort, sd, group=cohort, fill=cohort)) + 
    geom_jitter(width=0.1) + geom_violin(alpha=0.5) + theme(legend.position = 'none') +
    labs(title='SD'),
  ncol=2, top='Per bin scaled tile values', bottom='Val glob = validation bins normalized using global z-scores')
```

```{r}
preds = do.call(bind_rows, purrr::map(
  list.files('~/Data/BarrettsProgressionRisk/Analysis/validation/multipcf/predictions_5e6_all_var/0.9/', pattern='predictions.tsv', recursive = T, full.names = T), 
  function(f) read_tsv(f, col_types = 'cddcccccccc'))) %>% mutate(Endoscopy = as.Date(Endoscopy)) %>% dplyr::select(-`Path Notes`) %>% mutate(`Hospital Research ID` = str_replace_all( str_remove_all(`Hospital Research ID`, " "), '/', '_'))

@@

# Model trained on all samples

In this case I'm looking at the leave-one-out predictions, so somewhat different but at least gives an idea. While the progressors don't change, the non-progressors see a signficant shift.

load('~/Data/BarrettsProgressionRisk/Analysis/all-samples/loo.Rdata')

pg.samp = pg.samp %>% filter(Hospital.Research.ID %in% all.val$`Hospital Research ID` & !grepl('OCCAMS',Hospital.Research.ID)) %>%
  mutate(qt = cut(Prediction, breaks=seq(0,1,.1)))

preds = pg.samp %>% left_join(BarrettsProgressionRisk:::pred.confidence %>% dplyr::select(quant,Risk), by = c('qt'='quant'))

# Progressors
# FNR is still over 50% in the old progressors as well
fnr = round(nrow(preds %>% filter(Status == 'P' & Risk == 'Low' & !grepl('OCCAMS',`Hospital.Research.ID`)))/nrow(preds %>% filter(Status == 'P' &  !grepl('OCCAMS',`Hospital.Research.ID`))),2)

preds %>% dplyr::filter(Status == 'P' & !grepl('OCCAMS',`Hospital.Research.ID`)) %>% group_by(Risk) %>% tally() %>% spread(Risk, n) %>% 
  kable(caption=paste0('P risk predictions, FNR=',fnr)) %>% kable_styling(full_width = F) 
```
